---
title: "K-mer matching probabilities"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 3
  pdf_document:
    highlight: zenburn
    toc: yes
    toc_depth: 3
css: ../../html/course.css
---

## Parameters

```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE, comment=''}
## Define the knitr options
library(knitr)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment = "")
```

```{r parameters}
## Nucleotide matching probability 
## (assuming equiprobability and independence !)
p <- 1/4  ## prior residue probability
G <- 3e9 ## Genome size
k <- 26  ## Read length
```


## Concepts

### Bernoulli process

A Bernoulli process is defined as a succession of trials, where 

1. each trial can result in two possible (and exclusive) outomes: success or failure;
2. successive trials are independent from each oter;
2. the probability of success ($p$) is constant. 




## Probability to observe a succession of $k-1$ matches  followed by 1 mismatch

The geometric distribution describes the probability to observe $k-1$ successes followed by 1 failure, in a Bernoulli process. 

```{r geometric distribution}
x <- 1:30
dens.geo <- p^(x-1) * (1-p)
plot(x, dens.geo, type="h", lwd=3, col="blue",
     main="Geometric distribution",
     xlab="Read length", ylab="P(k-1 matches then 1 mm)")

```

### Etape XX

### Etape XX

### Etape XX

### Etape XX

### Etape XX

 
 
 ```{r choose}
plot(0:k, choose(k=0:k, n=k), lwd=3, col="green", type="h",
main="Choose", xlab="k", ylab="Number of combinations")
```

## Etape 6: probability to observer exactly $x$ matches and $k-x$ mismatches, at any position

$P(X=x) = C^x_k \cdot p^x \cdot (1-p)^{k-x}$

```{r binomial}
X <- 0:26

p.matches.eq.x <- dbinom(x=X, prob = p, size = k)

plot(X, p.matches.eq.x, type="h", lwd=3, col="cyan",
     
     main="Binomial density", xlab="Number of matches", ylab="P(X=x)")

exp.x <- k*p
abline(v=exp.x, col="darkgreen")

## A more general definition of the expectation: the sum of values weighted by their priorprobability
exp.x <- sum(X*p.matches.eq.x)

```

